# GCP setup information for instructors

```{r setup-gcp, echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

*Note*: This protocol assumes that you're using the Hallam-Lab Acount.

## Bucket Setup with the following parameters

1.  Click "Create a Bucket".
2.  Name the bucket
3.  Set Location to us-west1-b, Oregon
4.  Set storage class to standard
5.  Set Access Controls to Uniform
6.  Set Public Access to Not Public
7.  Now you can enter the bucket and create new directories if you wish
8.  Download data SI_72 data from Sharknado and upload them by dragging and dropping them into the Bucket (faster to upload one data type at a time, i.e. `MetaG_Assemblies`, `MetaG_Trimmed_Reads`, `MetaT_Trimmed_Reads`)

## Student Account Set-Up

1.  Click Add People To This Project under the Project Info Card. Roles can be changed with in the IAM card with the master account.

2.  Enter email and assign the following roles:

    a.  Compute Instance Admin (beta)
    b.  Compute Network User
    c.  Service Account User
    d.  Storage Object Viewer

## VM Set-Up

1.  Make VM on GCP using template:

    a.  Name the VM
    b.  Set Location to us-west1-b, Oregon
    c.  Change OS to Ubuntu 18.04
    d.  Set memory to 100 GB

2.  Login to instance with `gcloud compute ssh <vm name>`

3.  In the VM, enter the following to make the directory that you'll mount the Bucket to:

```{bash}
cd /usr/local
sudo mkdir data
cd data
sudo mkdir MICB425
```

4.  Authenticate and set credentials:

```{bash}
gcloud auth application-default login
```

5.  Now it's time to mount the Bucket by following [`gsfuse` install instructions](https://github.com/GoogleCloudPlatform/gcsfuse/blob/master/docs/installing.md)

6.  Logout of the VM and go back to compute engine page. Click on VM name, then edit. Go to metadata and add: `enable-oslogin  TRUE`

7.  Login to VM again and mount the bucket to the directory you created earlier:

```{bash}
cd /usr/local/data
sudo gcsfuse -o allow_other --implicit-dirs micb425-treesapp-data MICB425
```

Note:

-   `-o allow_other` lets all users access the bucket and override privacy permissions to do so.
-   `--implicit-dirs` gets all sub-directories and files in the bucket mounted to the VM.
-   To Dismount Bucket: `sudo fusermount -u MICB425` (or use directory name).

8.  To make sure everything is accessible, try to navigate through the bucket and see if you can get the first few lines of a MetaG assembly file.

## Setting up TreeSAPP singularity container

1.  Follow [Singularity Quick Start Install](https://sylabs.io/guides/3.5/user-guide/quick_start.html#quick-installation-steps) by copying the provided commands provided into the VM terminal.

2.  Download the TreeSAPP Container and rename and move it so it is accessible to all users:

```{bash}
singularity pull library://cmorganl/default/treesapp:sha256.f4d9901467c3a99c05a997175e71897dbcd38cd22d6d9bb9b79a86c1a66e1964
sudo mv treesapp_sha256.f4d9901467c3a99c05a997175e71897dbcd38cd22d6d9bb9b79a86c1a66e1964.sif /usr/local/bin/treesapp.sif
```

3.  Test container:

```{bash}
singularity exec /usr/local/bin/treesapp.sif treesapp info
```

```{r eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
# Reset code chunk settings to `default_chunk_options` (defined in `index.Rmd`)
knitr::opts_chunk$set(default_chunk_options)
```
