# Basic Automation with R

## Introduction and goals

The library `dplyr` provides many useful functions like `select()`, `filter()`, and `mutate()` that cover the vast majority of use cases, but real data is messy. Sometimes you may what to do but not *how* to do it with existing functions. This tutorial will teach you some fundamental building blocks of scripting so that you can produce the "how" when you know the "what."

A warning though, if using `dplyr` is driving a car, then this tutorial is teaching you how to walk. While more locations are reachable by walking, you will always get there much faster by driving. *Try to find the tools first before hacking together your own.*

## Prerequisites

```{r}
library(tidyverse)
```

## Making a function

In math, we specify functions like so: $f(x) = 2x$, where the function $f$ takes the parameter $x$ and performs the action $2x$ before handing back the result. In R, the same would look like this...

```{r}
f <- function(x) {
  return(2 * x)
}
```

...and it can be used like so.

```{r}
f(2)
```

## Decisions with conditionals

`if` allows selective execution based on a condition.

```{r}
doubleIfLessThan <- function(x, threshold) {
  if (x < threshold) {
    return(2*x)
  } else {
    return(x) # if no condition satisfied
  }
}

print(doubleIfLessThan(12, 100)) # 12 < 100, so doubled 12
print(doubleIfLessThan(12, 10)) # 12 ≮ 10, so just give back 12
```

More than one condition can be used...

```{r}
doubleOutsideRange <- function(x, lower, upper) {
  if (x < lower) {
    return(2*x)
  } else if (x > upper) {
    return(2*x)
  } else {
    return(x)
  }
}

print(doubleOutsideRange(5, 10, 20))
print(doubleOutsideRange(5, 0, 2))
print(doubleOutsideRange(5, 0, 10))
```

...and conditions can be combined with logical operators

```{r}
doubleIfInRange <- function(x, lower, upper) {
  if (x >= lower && x <= upper) { # && is the same as AND
    return(2*x)
  } else {
    return(x)
  }
}

print(doubleIfInRange(5, 10, 20))
print(doubleIfInRange(5, 0, 2))
print(doubleIfInRange(5, 0, 10))
```

| Operator |                                  Description                                   |
|:--------:|:------------------------------------------------------------------------------:|
|   `&&`   |                         AND, both sides must be `TRUE`                         |
|   `||`   |                      OR, at lease one side must be`TRUE`                       |
|   `!`    | NOT, inverts the following (`TRUE` becomes `FALSE` and `FALSE` becomes `TRUE`) |
|   `==`   |                         EQUALS, left must equal right                          |

Try the following

```{r}
1==1
1==1 && 2==1
!(2==1) || FALSE
```

## Iterating with loops

Suppose we wanted to double each number in a list.

```{r}
numbers <- read.csv(file="data/numbers.csv", header=TRUE)
numbers
```

We can use a `for` loop. This is how it behaves...

```{r}
for (i in 1:5) {
  print(i)
}
```

...and this is how we to use it in our context

```{r}
for (rowNumber in 1:nrow(numbers)) {
  n <- numbers[rowNumber, "n"] # get value from row "rowNumber", column "n"
  print(2 * n)
}
```

If the number of repeats is unknown, we can use a `while` loop. A condition is used instead to tell it when to stop.

```{r}
n <- 1
while (n < 1000) {
  n = n * 2
}

n
```

Loops can repeat forever. Click the red octagon (🛑) at the top right of your RStudio console to force the script to stop.

```{r}
while (TRUE) {
  # loop forever
}
```

## Reading files

Files can be opened and read one line at a time like so.

```{r}
numbersFileConnection=file("data/numbers.csv", "r") # r is for read

while (TRUE) {
  line = readLines(numbersFileConnection, n = 1) # read 1 line
  if (length(line) == 0) { # if no more lines ...
    break # ... exit loop
  }
  print(line)
}

close(numbersFileConnection)
```

The last parameter specifies the read/write permissions of the opened file.

```{r, error=TRUE}
con=file("data/numbers.csv", "r")
write("some new text", file=con)
```

Files can also be written to one line at a time like so.

```{r}
con=file("data/myNumbers.csv", "w")

write("n", con) # header
for (i in 1:10) {
  line <- toString(i+10)
  write(line, con)
}

close(con)
```

Check your `data` folder for the new file.

Reading and writing `csvs` are done *much better* by `read.csv()` and `write.csv()`, but suppose you come across this file...

```{r}
con=file("data/brokenNumbers.csv","r")
line=readLines(con) 
for (i in 1:10) {
  print(line[i])
}
close(con)
```

...seems innocent enough, but when trying to load it as a csv...

```{r}
isOdd <- read.csv(file="data/brokenNumbers.csv", header=TRUE)
isOdd %>% slice(50:60)
```

...it's format is obviously broken. Investigate the file and see if you can fix it.
[strsplit](https://stringr.tidyverse.org/reference/str_split.html)

```{r}
con=file("data/brokenNumbers.csv", "r")
outCon=file("data/fixedNumbers.csv", "w")

headers <- readLines(con, n = 1)
while (TRUE) {
  row = readLines(con, n = 1)
  if (length(row) == 0) {
    break
  }
  
  fixedRow <- ""
  
  # split row into cells
  cells <- str_split(row, ", ")[[1]] # see str_split above
  
  # look at cells by index like so
  twoN = cells[length(cells)]
  
  # fixedRow = ...

  write(fixedRow, outCon)
}

close(con)
close(outCon)
```
