# Analysing the TreeSAPP classifications in R

```{r setup-R, echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Introduction and Objectives

These instructions will help you to complete the script template `treesapp_analysis.R` to complete the initial analysis of your TreeSAPP classifications.
The following code chunks will show you how to load, aggregate, subset and transform continuous and discrete variables in R using the `tidyverse`.
You will also be provided with examples for plotting TreeSAPP classification data using the `ggplot2` library.

## Resources for analysis of TreeSAPP output in R

-   The TreeSAPP output file `extra_annotated_marker_contig_map.tsv` with the metagenome and metatranscriptome abundances and gene functions you generated with [TreeSAPP](#running-treesapp on a server).
-   An R script template called `treesapp_analysis.R` on Canvas.
-   `Saanich_TimeSeries_Chemical_DATA.csv` file containing geochemical measurements.

## Checklist to write and run R script

- [ ] Place `treesapp_analysis.R` and both `extra_annotated_marker_contig_map.tsv` files into a single folder and create a new RStudio project in that folder on your local computer.
- [ ] Edit the `treesapp_analysis.R` script following [the instructions below](#writing-the-r-script).

## Writing the R script

In the `treesapp_analysis.R` script you will load and subset your TreeSAPP data to the variables and marker genes of interest. You will then break the one taxonomic information column into multiple taxonomic ranks. You will also load the `Saanich_TimeSeries_Chemical_DATA.csv` file for geochemical measurements to learn more about the conditions at your assigned depth.
Fianlly, you will 

## Load any required packages

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Load metagenomic and metatranscriptomic TreeSAPP data

```{r}
ts_dat <- read.table(file="data/extra_annotated_marker_contig_map.tsv",
                     header=TRUE,
                     sep="\t", quote="")
```

### Classification table

Let's look at TreeSAPP's classification table for these FunGene XmoA sequences. In RStudio, you are able to click on a variable name under the "Environment" tab in the top-right panel. Optionally, you can use the `View` function in the R-console to view an object. For example, you could run `View(ts_dat)`.

A description of each of the fields can be found under the [`treesapp assign` documentation page](https://github.com/hallamlab/TreeSAPP/wiki/Classifying-sequences-with-treesapp-assign#classification-table).

## Subset your data to the variables and marker genes of interest.

Determine which variables and marker genes you will need and subset your data to them. The only "Marker" in these data is XmoA, so we will filter to XmoA.

```{r}
ts_df <- ts_dat %>%
  filter(Marker == "XmoA")
```

When working with real data you may be analyzing multiple reference package's worth of data. If you want to subset those data to a couple of reference packages you will find the '%in%' operator handy. In the following example we are keeping data where the Marker variable is equal to any of "XmoA", "PmoA" or "AmoA". In reality, they will only match "XmoA".

```{r}
ts_df <- ts_df %>%
  filter(Marker %in% c("XmoA", "PmoA", "AmoA"))
```

## Reformat variables

### Processing taxonomic information

Your taxonomic information contains information from the highest to lowest available rank. Look at a few rows containing taxonomic information and determine the separator between each rank. We can use a function in the tidyverse called `separate` that allows you split a character object at each of these separators. As a result, your initial single taxonomic variable will be split into multiple new variables, one for each rank. Use the R object `taxa_ranks` to assign the names for those new variables.

```{r}
taxa_ranks <- c("Root", "Domain", "Phylum", "Class",
                "Order", "Family", "Genus", "Species")
ts_df <- ts_df %>%
  separate(col = Taxonomy,
           into = taxa_ranks,
           sep = "; ", fill = "right", remove = T)
```

### Distinguishing metagenomes from metatranscriptomes

```{r}
ts_df <- ts_df %>% 
  mutate(Sample = gsub("_pe", "_MetaG", Sample)) %>% 
  separate(col = Sample,
           into = c("Cruise", "Depth", "SeqType"),
           extra = "drop")
```

### Processing depth data

Modify the data frame to split the sample name into "Cruise" and "Depth" variables. Then, remove the 'm' character from the "Depth" variable to create a numeric variable called "Depth.m".

```{r}
ts_df <- ts_df %>%
  mutate(Depth.m = as.numeric(gsub('m', '', Depth))) %>% 
  mutate(Cruise = as.numeric(gsub('SI0', '', Cruise)))
```

## Load the geochemical data

Geochemical data for the Saanich Inlet time series is in another file called `Saanich_TimeSeries_Chemical_DATA.csv`. We will first load it into its own data frame called `geochem_df`, and subset it to just the samples we are interested in.

```{r}
geochem_df <- read.table("data/Saanich_TimeSeries_Chemical_DATA.csv",
                          header=TRUE, sep=',') %>% 
  filter(Cruise == 72) %>% 
  select(Cruise, Depth,
         CTD_O2, NO3,
         Mean_NH4, Mean_NO2, Mean_H2S, Mean_CH4) %>% 
  mutate(Mean_CH4 = Mean_CH4*1E-3) %>% 
  mutate(across(.fns=as.numeric))
```

We are going to use the function `left_join` from the `dplyr` library to merge the geochemical data for cruise 72 with the TreeSAPP classifications for XmoA by Sample. The combined data will be saved to a new data frame called `ts_geo_df`.

```{r}
ts_geo_df <- left_join(ts_df,
                       geochem_df,
                       by=c("Cruise",
                            "Depth.m" = "Depth"))
```

## Visualize the taxonomic, functional and chemical data from Saanich Inlet

### Analyze TreeSAPP's taxonomic classifications

Distribution of abundances of XmoA taxonomic orders  at different depths

```{r}
ts_geo_df %>%
  group_by(Depth.m, Order) %>%
  summarise(sum = sum(Abundance)) %>%
  ungroup() %>%
  ggplot(aes(x=Depth.m, y=sum, colour=Order)) +
  coord_flip() +
  scale_x_reverse() +
  geom_line() +
  labs(x="Depth (m)",
       y="Relative abundance (FPKM)")
```

Barplot demonstrates relative change in proportions at different depths

```{r}
ts_geo_df %>% 
  group_by(Depth, Depth.m) %>% 
  mutate(Proportion = Abundance/sum(Abundance)) %>% 
  group_by(Depth, Depth.m, Order) %>% 
  summarise(sum = sum(Proportion)) %>% 
  ungroup() %>% 
  mutate(Depth = reorder(Depth, Depth.m)) %>% 
  ggplot(aes(x=Depth, y=sum, fill=Order)) +
  geom_bar(stat = "identity") +
  labs(x="Depth",
       y="Relative abundance")

```

### Analyze the functional data

```{r}
ts_geo_df %>% 
  group_by(Depth, Depth.m, SeqType, Order, Function) %>% 
  summarise(Sum = sum(Abundance)) %>% 
  ungroup() %>%
  mutate(Depth = reorder(Depth, desc(Depth.m))) %>% 
  ggplot(aes(x=Depth, y=Order, colour=Function)) +
  coord_flip() +
  facet_wrap(~SeqType) +
  geom_point(aes(size=Sum)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Analyze the chemical measurements 

```{r}
ts_geo_df %>% 
  pivot_longer(cols=c(starts_with("Mean"), NO3, CTD_O2),
               values_to = "Value.uM",
               names_to = "Molecule") %>%
  ggplot(aes(x=Depth.m, y=Value.uM)) +
  geom_point() +
  geom_line() +
  facet_wrap(~Molecule, scales = "free_x") +
  coord_flip() +
  scale_x_reverse()
```

### Bringing them all together

```{r}
ts_geo_df %>%
  filter(SeqType=="MetaT") %>%
  filter(!is.na(Order)) %>% 
  mutate(Depth = reorder(Depth, Depth.m)) %>% 
  ggplot(aes(x=Depth, y=Abundance, colour=Mean_CH4)) +
  facet_wrap(~Order) +
  geom_boxplot() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Analyzing classifications for your group's reference package

- How does abundance of your gene vary with depth? Are trends similar for both RNA and DNA?

- How does microbial diversity vary with depth? Are trends similar for both RNA and DNA?
You will need to make a decision at what taxonomic rank you will calculate diversity. Why might you not be able to use the lowest possible? Why should your taxonomic rank not be too high?

- How does the abundance of your gene relate to water column geochemistry in Saanich Inlet (use the chemical measurement data in `Saanich_TimeSeries_Chemical_DATA.csv`)?

- Determine the geochemical conditions at your given depths. Which is the best available terminal electron acceptor?

```{r eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
# Reset code chunk settings to `default_chunk_options` (defined in `index.Rmd`)
# knitr::opts_chunk$set(default_chunk_options)
```
